One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2024-11-12 paper.tex
update acknowledgements
H + R comments

The estimation of the abundance and spatial distribution of wild populations of animals is a critical objective in ecology and conservation \citep{schwarz_estimating_1999}.This paper presents a statistical analysis of survey data collected on a critically endangered Hawaiian forest bird, offering insights into its population size and spatial dynamics, which crucial to guide effective conservation management decisions.  The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.  The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution through time.

@@

Since the existing \akepa{} population numbers in the thousands and lives in dense tropical forest where logistical and ecological challenges mean a full census of the population is not feasible. The \hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986} is a large-scale, quantitative survey of Hawaiian forest birds.  Annual surveys of the \akepa{} study-region consist of a number of point transects located at approximately regular intervals along randomly located line transects.  At these surveyed locations, the detectability of animals is unknown and must be estimated in order to convert observed counts into estimates of the total population size.  The \akepa{} survey estimates detectability using a point transect distance sampling approach \citep{buckland_distance_2015} where, for each observation, the distance to the observer is recorded.

Additionally, a key aim of the analysis is to predict animal density at unsurveyed locations.  If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed location.  However, in many ecological contexts there is good reason to believe there are drivers of the spatial distribution for which we have no suitable explanatory covariates available. For the \akepa{} data, there is a clear example of an unexplained driver of the spatial distribution where the population favours the southern region of the preserve.  No straightforward explanatory cause has been found for this preference \citep{camp_dsm_2020}.  From a statistical perspective, we address this unexplained heterogeneity by including a spatially structured random effect in the model.

@@

The paper is structured as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the spatially-structured random effect; and Sections \ref{sec-results} presents results of the analysis, model evaluation and effective communication of uncertainty.

@@

The \hawaii{} \akepa{} is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.  Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the HFBS \citep{scott_HFBS_1986}. Information from the HFBS is used to update the International Union for Conservation of Nature redlist \citep{iucn_redlist_2022}, where the \akepa{} is currently listed as endangered. HFBS data is also used to establish preserves that coincide with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau).  Hakalau is the first wildlife refuge established in \hawaii{} with the primary purpose to protect and manage native forests for the conservation of threatened and endangered bird and plant species.

@@

Hakalau was established in 1985 to conserve 15,390 ha of montane forest habitat for native forest birds and rainforest plants. Annual forest bird surveys were initiated in 1987 to determine the population status and track trends in abundance. Survey points were established along 14 line transects following a systematic, random design with point transects located approximately 150m apart on line transects located either 500m or 1,000m apart. We limit our study area to the open-forest and closed-forest strata of Hakalau (\autoref{fig:2002studyareapointspt}), an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016}, who omitted the closed-forest stratum because it was not sampled in the early years of the surveys.  For our analysis, we use data from a later year in which the closed-forest stratum was sampled and so we are able to include it.  The open-forest stratum was previously heavily grazed, and, since the removal of cattle in 1988, regeneration has proceeded naturally \citep{maxfield_hakalau_1998}. The closed-forest stratum was historically least modified by grazing and is relatively intact forest habitat.  The northern boundary of the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig. \ref{fig:2002studyareapointspt}). The southern boundary is the same as that in \cite{camp_population_2010}, chosen to exclude unsurveyed regions to the south of Hakalau but does not represent a physical boundary. To the west of the study area is pasture that is dominated by grass and is unsuitable habitat for \akepa{} and the boundary here marks the edge of the forest.

\begin{figure}[!htb]
	\centering
	\includegraphics{figures/study_area_design.pdf}
	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.  The red outline marks the study area boundary as defined in \cite{camp_dsm_2020}.}
	\label{fig:2002studyareapointspt}
\end{figure}

The survey follows a point-transect distance sampling design, recording the horizontal distances from observers to detected birds. Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength greater than Beaufort scale 3). During 8-minute counts, trained observers recorded the species, distance to the nearest metre for each bird detected. \cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the study area and the bird surveys.

To demonstrate our approach we select a single survey year from the \akepa{} time series that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability. In 2002, there were a total of 289 point transects sampled within the 4,603 ha study area of Hakalau (Fig. \ref{fig:2002studyareapointspt}).  A total of 276 \akepa{} were detected on 121 point transects. The number of detections within each point transect ranged from zero to six. We selected data from a single year to demonstrate our approach in a simplified setting that does not require a temporal model component.  The choice of year is somewhat arbitrary and our model has also been applied to data from other years with similar results to those presented here.  Our approach is readily extendible to a multi-year analysis, similar to that in \cite{camp_dsm_2020}, which we discuss in more detail in Section \ref{sec-discussion}.

@@

Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.  A popular one-stage maximum-likelihood method, introduced by \cite{royle_ModelingAbundanceEffects_2004} and implemented in the \texttt{R} package \texttt{unmarked} \citep{fiske_UnmarkedPackageFitting_2011}, is based on reformulating the model as a multinomial likelihood.  This depends on choosing discrete distance classes and binning the data into counts defined on discrete spatial units.  Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, the authors derive a Poisson likelihood to which standard methods maximum-likelihood methods can be applied. This approach rests heavily on the discretisation of both the distance data and the spatial location data.

@@

The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use INLA \citep{rue_approximate_2009} and present an application to line transect distance sampling data.  \citet{yuan_point_2017} also take a point process perspective and formulate the detection model as a thinning of a point pattern.  Key to their approach is formulating the detection model as the solution to an SPDE, a formulation that is unfamiliar to most distance sampling practitioners.

A downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, which is usually a key feature of detection functions. The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful. These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA for spatial modelling. Here, we present an approach that is inspired by \citet{yuan_point_2017} but with some key differences. We also fit the model using INLA and take advantage of the benefits of the increased computational efficiency compared to MCMC.  However, we use a well-known parametric family of detection functions, that is familiar to users of classical distance sampling methods. This results in components of the model predictor that are non-linear in their parameters, making model fitting infeasible using standard INLA methods \citep{rue_approximate_2009}.  We address this by using a novel method of iterated model fits based on a first-order Taylor expansion of the non-linear model components.  Our approach is implemented in the \texttt{inlabru} package \citep{lindgren_inlabru_2024, bachl_inlabru_2019}, which also has support for spatial data objects.

@@

To describe the model in general we assume a point transect distance sampling survey consists of a set of $K$ point transects.   We denote $k$-th transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.  In the HFBS the point transects are non-overlapping disks all with with equal radius $W$ and centres at locations $\bs_k$ for $k = 1, \ldots, K$.  The probability of observing a point at location $\bs \in \Omega_k$, given there is a point at that location, we denote by $g_k(\bs)$.  The probability of observing a point outside the surveyed region is zero.
Since the point transects are non-overlapping, each location $\bs \in \Omega$ is unambiguously associated with a single observer and, therefore, a single thinning probability function $g_k$.   The half-normal detection function for points in $\Omega_k$ is therefore $g_k(\bs) = \exp(-\lVert \bs - \bs_k \rVert_2^2 / 2\sigma^2)$.  The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation and observer indentifiers.  The thinning probability function for any $\bs \in \Omega$ is then given by $g(\bs) = g_{k(\bs)}(\bs)$ where $k(\bs)$ is an indexing function such that $k(\bs) = k$ whenever $\bs \in \Omega_k$.

@@

To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.  In a slight abuse of notation $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes a location within the $k$-th point transect. The thinning function $g_k(\bs_k(r, \theta))$ depends only on $r$ and not on $\theta$ and so we use the shorthand notation $ g(\bs_k(r, \theta)) = g_k(r)$. The integral in the likelihood can be simplified using an assumption that $\lambda(\bs)$ is linear within each sampling unit.  This implies that $\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi)) = 2\lambda(\bs_k)$.  Therefore
\begin{align}
\label{eq-linear-intensity}
	\int_0^{2\pi} \lambda(\bs_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \{\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi) \} g_k(r)\mathrm{d}\theta \nonumber \\
	&= 2\pi \lambda(\bs_k)g_k(r).
\end{align}
We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.  A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piecewise linear basis functions used for the spatially-structured random effect described in Section \ref{sec-gmrf}.

@@

The family of models that can be fitted using the INLA methodology are known as latent Gaussian models.  These depend on a latent Gaussian vector of parameters which we denote $\bm{u}$. Parameters are related to observations via a predictor $\eta(\bm{u})$ which, for the LGCP model, is the log-intensity evaluated at the observed point locations and the integration scheme locations as described in Section \ref{sec-lgcp}.  Since $\phi$ is an element of $\bm{u}$, $\eta(\bm{u})$ is a non-linear function with respect to $\bm{u}$.

@@

The `optimal' linearisation point is defined via a fixed point iteration scheme. Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.  A new linearisation point is found by via the functional
\begin{equation}
	f(\ol{p}_{\bm{v}}) = (\wh{\bm{\theta}}_{\bm{v}},\wh{\bm{u}}_{\bm{v}})
\end{equation}
where
\begin{equation}
\wh{\bm{\theta}}_{\bm{v}} = \argmax_{\bm{\theta}} \ol{p}_{\bm{v}} ( \bm{\theta} | \bm{y}),
\end{equation}
the posterior mode for $\bm{\theta}$, and
\begin{equation}
\wh{\bm{u}}_{\bm{v}} = \argmax_{\bm{u}} \ol{p}_{\bm{v}} (\bm{u} | \bm{y}, \wh{\bm{\theta}}_{\bm{v}}),
\end{equation}
the joint conditional posterior mode for $\bm{u}$.  In other words, this functional defines a new linearisation point as the posterior mode for $\bm{\theta}$ under the linearised model configuration and the conditional mode for $\bm{u}$ given $\bm{\theta}$.  These posteriors are computed automatically during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost.

@@

We define a finite element mesh with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$. The GRF is represented as $\xi(\bs) = \sum_l \beta_l \phi_l(\bs)$.  The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices (\cite{blangiardo_spatial_2013}, Equation 6.19). See \cite{lindgren_explicit_2011} for full details.

@@

All figures that summarise the posterior intensity field have potential weaknesses that should be considered when communicating with stakeholders.  This sections discusses these weaknesses and presents and example of one approach that tries to address them.  The aim of this section is to show that our approach is readily extendible to produce bespoke model outputs.

@@

The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.  However, CV values will generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.  Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.  However, there is also variable sampling effort within the study region, with higher CV values in the region with lower effort.  It is not clear what impact this has on the CV values and this makes the map hard to interpret, even for statisticians who understand the model well. The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values also intended to indicate uncertainty.  However, a default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.  Whether these relative differences matter is dependent on the overall context of the analysis.  Some levels of variability may be acceptable in some contexts and unacceptable in others.

Quantile maps (Figure \ref{fig:intensity-quantiles}) are also difficult to interpret. The temptation is to perceive these maps as showing possible intensity surfaces that could have produced the observed data.  Stake holders may interpret the 0.025 quantile map as a kind of lower bound intensity surfaces and the 0.975 quantile map as an upper bound. However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.  A similar problem holds for maps of the marginal probability of exceeding a threshold at each prediction location. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics{figures/intensity_quantiles.pdf}
		\caption{Pointwise posterior quantiles for the intensity:  A) 0.025 quantile B) 0.975 quantile}
		\label{fig:intensity-quantiles}
	\end{center}
\end{figure}
These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.  To demonstrate the misleading nature of these maps, if we (incorrectly) treat the lower and upper quantile plots as though they are intensity surfaces and integrate them, then we obtain an abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.  However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).

Broadly speaking, we advocate for moving beyond maps of uncertainty presented for their own sake and towards model outputs that directly relate to the wider objectives of the analysis. We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult or impossible to interpret. In the following section we suggest an approach that avoids some, but not all, of the pitfalls above.

@@

\autoref{fig:excursions}A shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95 (hence $\alpha = 0.05$).  Note that we use this threshold as the definition of a hotspot, and proceed with generating model results that incorporate this definition.  This figure can be interpreted in a natural way as the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95. 
\begin{figure}[!htb]
	\includegraphics{figures/excursions.pdf}
	\caption{A) The blue area is the positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.  B) The positive excursion function with a level corresponding to 1 bird per hectare.  The scale shows the probability values $1-\alpha$.}
	\label{fig:excursions}
\end{figure} 
This map could be used to define, for example, a `core region' for the \akepa{} population, with particular conservation importance.  Crucially, this map depends on a clear mathematical description of what a `core region' actually means in a statistical sense.  While any particular definition would be up for debate, the advantage is that the statistical methodology is taken as close as possible to the aims of the analysis.  Often a map of species density will be interpreted informally in many different ways (one of which is identifying hotspots).  To visualise how such excursion sets change with different levels of uncertainty we can use the excursion function $F_u^{+}(\bs) = \sup \{1 - \alpha ; \bs \in E_{u,\alpha}^+ \}$.  This defines, for each location, the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.  I.e.  if we allow greater uncertainty, this function shows which locations would be included in the excursion set.

>>>>

The estimation of the abundance and spatial distribution of wild populations of animals is a critical objective in ecology and conservation \citep{schwarz_estimating_1999}.This paper presents a statistical analysis of survey data collected on a critically endangered Hawaiian forest bird, offering insights into its population size and spatial dynamics, which are crucial to guide effective conservation management decisions.  The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.  The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution through time.

@@

Since the existing \akepa{} population numbers in the thousands and lives in dense tropical forest, logistical and ecological challenges mean a full census of the population is not feasible. The \hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986} is a large-scale, quantitative survey of Hawaiian forest birds.  Annual surveys of the \akepa{} study-region consist of a number of point transects located at approximately regular intervals along randomly located line transects.  At these surveyed locations, the detectability of animals is unknown and must be estimated in order to convert observed counts into estimates of the total population size.  The \akepa{} survey estimates detectability using a point transect distance sampling approach \citep{buckland_distance_2015} where, for each observation, the distance to the observer is recorded.

Additionally, a key aim of the analysis is to predict animal density at unsurveyed locations.  If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed locations.  However, in many ecological contexts there is good reason to believe there are drivers of the spatial distribution for which we have no suitable explanatory covariates available. For the \akepa{} data, there is a clear example of an unexplained driver of the spatial distribution where the population favours the southern region of the preserve.  No straightforward explanatory cause has been found for this preference \citep{camp_dsm_2020}.  From a statistical perspective, we address this unexplained heterogeneity by including a spatially structured random effect in the model.

@@

The paper is structured as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} describes the iterated INLA fitting procedure and the spatially-structured random effect; and Sections \ref{sec-results} presents results of the analysis, model evaluation and effective communication of uncertainty.  Section \ref{sec-discussion} discusses potential extensions of the model and wider uses for iterative INLA.

@@

The \hawaii{} \akepa{} is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to the island of \hawaii{}, USA.  Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the HFBS \citep{scott_HFBS_1986}. Information from the HFBS is used to update the International Union for Conservation of Nature redlist \citep{iucn_redlist_2022}, where the \akepa{} is currently listed as endangered. HFBS data is also used to establish preserves that coincide with native bird hotspots, including Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{} (hereafter Hakalau).  Hakalau is the first wildlife refuge established in \hawaii{} with the primary purpose to protect and manage native forests for the conservation of threatened and endangered bird and plant species.

@@

Hakalau was established in 1985 to conserve 15,390 ha of montane forest habitat for native forest birds and rainforest plants. Annual forest bird surveys were initiated in 1987 to determine the population status and track trends in abundance. Survey points were established along 14 line transects following a systematic, random design with point transects located approximately 150m apart on line transects located either 500m or 1,000m apart. We limit our study area to the open-forest and closed-forest strata of Hakalau (\autoref{fig:2002studyareapointspt}), an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016}, who omitted the closed-forest stratum because it was not sampled in the early years of the surveys.  For our analysis, we use data from a later year in which the closed-forest stratum was sampled and so broaden the extent of spatial inference compared to previous studies.  The open-forest stratum was previously heavily grazed, and, since the removal of cattle in 1988, regeneration has proceeded naturally \citep{maxfield_hakalau_1998}. The closed-forest stratum was historically least modified by grazing and is relatively intact forest habitat.  The northern boundary of the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig. \ref{fig:2002studyareapointspt}). The southern boundary is the same as that in \cite{camp_population_2010}, chosen to exclude unsurveyed regions to the south of Hakalau but does not represent a physical boundary. To the west of the study area is pasture that is dominated by grass and is unsuitable habitat for \akepa{} and the boundary here marks the edge of the forest.

\begin{figure}[!htb]
	\centering
	\includegraphics{figures/study_area_design.pdf}
	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{}.  The red outline marks the study area boundary as defined in \cite{camp_dsm_2020}.}
	\label{fig:2002studyareapointspt}
\end{figure}

The survey follows a point transect distance sampling design, recording the horizontal distances from observers to detected birds. Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength greater than Beaufort scale 3). During 8-minute counts, trained observers recorded the species, distance to the nearest metre for each bird detected. \cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the study area and the bird surveys.

To demonstrate our approach we selected a single survey year from the \akepa{} time series that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability. In 2002, there were a total of 289 point transects sampled within the 4,603 ha study area of Hakalau (Fig. \ref{fig:2002studyareapointspt}).  A total of 276 \akepa{} was detected on 121 point transects. The number of detections within each point transect ranged from zero to six. We selected data from a single year to demonstrate our approach in a simplified setting that does not require a temporal model component.  The choice of year is somewhat arbitrary and our model has also been applied to data from other years with similar results to those presented here.  Our approach is readily extendible to a multi-year analysis, similar to that in \cite{camp_dsm_2020}, which we discuss in more detail in Section \ref{sec-discussion}.

@@

Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.  A popular one-stage maximum-likelihood method, introduced by \cite{royle_ModelingAbundanceEffects_2004} and implemented in the \texttt{R} package \texttt{unmarked} \citep{fiske_UnmarkedPackageFitting_2011}, is based on reformulating the model as a multinomial likelihood.  This depends on choosing discrete distance classes and binning the data into counts defined on discrete spatial units.  Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, the authors derive a Poisson likelihood to which standard maximum-likelihood methods can be applied. This approach rests heavily on the discretisation of both the distance data and the spatial location data.

@@

The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use INLA and present an application to line transect distance sampling data.  \citet{yuan_point_2017} also take a point process perspective and formulate the detection model as a thinning of a point pattern.  Key to their approach is formulating the detection model as the solution to an stochastic partial differential equation (SPDE), a formulation that is unfamiliar to most distance sampling practitioners.

A downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, which is a critical assumption of detection functions \citep{buckland_distance_2015}. The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful. These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA for spatial modelling. Here, we present an approach that is inspired by \citet{yuan_point_2017} but with some key differences. We also fit the model using INLA and take advantage of the benefits of the increased computational efficiency compared to MCMC.  However, we use a well-known parametric family of detection functions, that is familiar to users of classical distance sampling methods. This results in components of the model predictor that are non-linear in their parameters, making model fitting infeasible using standard INLA methods \citep{rue_approximate_2009}.  We address this by using a novel method of iterated model fits based on a first-order Taylor expansion of the non-linear model components.  Our approach is implemented in the \texttt{inlabru} package \citep{lindgren_inlabru_2024, bachl_inlabru_2019}, which also has support for spatial data objects.

@@

To describe the model in general we assume a point transect distance sampling survey consists of a set of $K$ point transects.   We denote $k$-th point transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.  In the HFBS the point transects are non-overlapping disks all with equal radius $W$ and centres at locations $\bs_k$ for $k = 1, \ldots, K$.  The probability of observing a point at location $\bs \in \Omega_k$, given there is a point at that location, we denote by $g_k(\bs)$.  The probability of observing a point outside the surveyed region is zero.
Since the point transects are non-overlapping, each location $\bs \in \Omega$ is unambiguously associated with a single observer and, therefore, a single thinning probability function $g_k$.   The half-normal detection function for points in $\Omega_k$ is therefore $g_k(\bs) = \exp(-\lVert \bs - \bs_k \rVert_2^2 / 2\sigma^2)$.  The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation and observer indentifiers.  The thinning probability function for any $\bs \in \Omega$ is then given by $g(\bs) = g_{k(\bs)}(\bs)$ where $k(\bs)$ is an indexing function such that $k(\bs) = k$ whenever $\bs \in \Omega_k$.

@@

To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.  In a slight abuse of notation, $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes a location within the $k$-th point transect. The thinning function $g_k(\bs_k(r, \theta))$ depends only on $r$ and not on $\theta$ and so we use the shorthand notation $ g(\bs_k(r, \theta)) = g_k(r)$. The integral in the likelihood can be simplified using an assumption that $\lambda(\bs)$ is linear within each sampling unit.  This implies that $\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi)) = 2\lambda(\bs_k)$.  Therefore
\begin{align}
\label{eq-linear-intensity}
	\int_0^{2\pi} \lambda(\bs_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \{\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi) \} g_k(r)\mathrm{d}\theta \nonumber \\
	&= 2\pi \lambda(\bs_k)g_k(r).
\end{align}
We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.  A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piecewise linear basis functions used for the spatially-structured random effect described in Section \ref{sec-gmrf}.

@@

The family of models that can be fitted using the INLA methodology are known as latent Gaussian models.  These depend on a latent Gaussian vector of parameters which we denote $\bm{u}$. Parameters are related to observations via a predictor $\eta(\bm{u})$ which, for the log-Gaussian Cox process model, is the log-intensity evaluated at the observed point locations and the integration scheme locations as described in Section \ref{sec-lgcp}.  Since $\phi$ is an element of $\bm{u}$, $\eta(\bm{u})$ is a non-linear function with respect to $\bm{u}$.

@@

The `optimal' linearisation point is defined via a fixed point iteration scheme. Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.  A new linearisation point is found via the functional
\begin{equation}
	f(\ol{p}_{\bm{v}}) = (\wh{\bm{\theta}}_{\bm{v}},\wh{\bm{u}}_{\bm{v}})
\end{equation}
where
\begin{equation}
\wh{\bm{\theta}}_{\bm{v}} = \argmax_{\bm{\theta}} \ol{p}_{\bm{v}} ( \bm{\theta} | \bm{y}),
\end{equation}
the posterior mode for $\bm{\theta}$, and
\begin{equation}
\wh{\bm{u}}_{\bm{v}} = \argmax_{\bm{u}} \ol{p}_{\bm{v}} (\bm{u} | \bm{y}, \wh{\bm{\theta}}_{\bm{v}}),
\end{equation}
the joint conditional posterior mode for $\bm{u}$.  In other words, this functional defines a new linearisation point as the posterior mode for $\bm{\theta}$ under the linearised model configuration and the conditional mode for $\bm{u}$ given $\bm{\theta}$.  These posteriors are computed automatically during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost.

@@

We define a finite element mesh with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$. The GRF is represented as $\xi(\bs) = \sum_l \beta_l \phi_l(\bs)$.  The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices that depend on a triangulation of the spatial domain (\cite{blangiardo_spatial_2013}, Equation 6.19). See \cite{lindgren_explicit_2011} for full details.

@@

All figures that summarise the posterior intensity field have potential weaknesses that should be considered when communicating with stakeholders.  This sections discusses these weaknesses and presents an example of one approach that tries to address them.  The aim of this section is to show that our approach is readily extendible to produce bespoke model outputs.

@@

The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.  However, CV values will generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.  Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.  However, there is also variable sampling effort within the study region, with higher CV values in the region with lower effort.  It is not clear what impact this has on the CV values and this makes the map hard to interpret, even for statisticians who understand the model well. The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values that are also intended to indicate uncertainty.  However, a default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.  Whether these relative differences matter is dependent on the overall context of the analysis.  Some levels of variability may be acceptable in some contexts and unacceptable in others.

Quantile maps (Figure \ref{fig:intensity-quantiles}) are also difficult to interpret. The temptation is to perceive these maps as showing possible intensity surfaces that could have produced the observed data.  Stakeholders may interpret the 0.025 quantile map as a kind of lower bound intensity surfaces and the 0.975 quantile map as an upper bound. However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.  A similar problem holds for maps of the marginal probability of exceeding a threshold at each prediction location. 
\begin{figure}[!htb]
	\begin{center}
		\includegraphics{figures/intensity_quantiles.pdf}
		\caption{Pointwise posterior quantiles for the intensity:  A) 0.025 quantile B) 0.975 quantile}
		\label{fig:intensity-quantiles}
	\end{center}
\end{figure}
These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.  To demonstrate the misleading nature of these maps, if we were to (incorrectly) treat the lower and upper quantile plots as though they were intensity surfaces and integrate them, then we would obtain an abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.  However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).  In other words, these maps tempt an interpretation that is inconsistent with the very model that produced the maps.

Broadly speaking, we advocate for moving beyond maps of uncertainty presented for their own sake and towards model outputs that directly relate to the wider objectives of the analysis. We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult to interpret. In the following section we suggest an approach that avoids some, but not all, of the pitfalls above.

@@

\autoref{fig:excursions}A shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95 (hence $\alpha = 0.05$).  Note that we use this threshold as the definition of a hotspot, and proceed with generating model results that incorporate this definition.  This figure can be interpreted in a natural way as the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95. 
\begin{figure}[!htb]
	\includegraphics{figures/excursions.pdf}
	\caption{A) The blue area is the positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.  B) The positive excursion function with a level corresponding to 1 bird per hectare.  The scale shows the probability values $1-\alpha$.}
	\label{fig:excursions}
\end{figure} 
This map could be used to define, for example, a `core region' for the \akepa{} population, with particular conservation importance.  Crucially, this map depends on a clear mathematical description of what a `core region' actually means in a statistical sense.  While any particular definition would be up for debate, the advantage is that the statistical methodology is taken as close as possible to the aims of the analysis.  Often a map of species density will be interpreted informally in many different ways (one of which is identifying hotspots).  To visualise how such excursion sets change with different levels of uncertainty we can use the excursion function $F_u^{+}(\bs) = \sup \{1 - \alpha ; \bs \in E_{u,\alpha}^+ \}$.  This defines, for each location, the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.  That is, if we allow greater uncertainty, this function shows which locations would be included in the excursion set.
